\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[margin=1.5cm]{geometry}

\begin{document}

\section{Notation and Definitions}

All definitions are tailored to single-gameweek FPL optimization: maximizing expected points under budget (£100m),
squad composition (15 players: 2 GK, 5 DEF, 5 MID, 3 FWD), club limits ($\leq$3 per club), lineup formation, and bench rules (1 GK, 3 outfield).

\subsection{Sets}
\begin{itemize}
    \item $P$: Set of all players.
    \item $C$: Set of clubs.
    \item $G \subset P$: Goalkeepers (GK). 
    \item $D \subset P$: Defenders (DEF).
    \item $M \subset P$: Midfielders (MID).
    \item $F \subset P$: Forwards (FWD).
    \item $S \subset P$: Set of starting XI players.
    \item $B \subset P$: Set of bench players.
    \item $Q = S \oplus B \subset P$: Set of all squad players.
\end{itemize}

\subsection{Parameters}
\begin{itemize}
    \item $q_i \in [0,1]$: No-show probability for player $i \in P$.
    \item $p_i \in \mathbb{R}$: Cost of player $i \in P$.
    \item $u_i \in \{0,1\}$: 1 if player $i \in P$ is unavailable for the upcoming gameweek, 0 if available.
    \item $y_i^0 \in \{0,1\}$: 1 if player $i$ is in current squad (known state, for Post-GW1).
    \item $B_{\mathrm{bank}} \in \mathbb{R}^+ \cup \{0\}$: Cash in bank (known state, for Post-GW1).
    \item $p^0_{\text{buy},i} \in \mathbb{R}^+ \cup \{0\}$: Purchase price paid for player $i \in P$ 
    \item $sp_i \in \mathbb{R}^+$: Selling price for player $i \in P$.
    \item $f \in \{0,1,2,3,4,5\}$: Free transfers (1 + banked, $\leq 5$).
    \item $\mathrm{expected\_points}_i = \mathrm{predicted\_points}_i \times (1 - q_i)$: Expected points for player $i \in P$ in upcoming gameweek.
\end{itemize}

\subsection{Variables}
\begin{itemize}
    \item $x_i \in \{0,1\}$: 1 if player $i \in P$ is in the starting XI .
    \item $c_i \in \{0,1\}$: 1 if player $i \in P$ is captain .
    \item $y_i \in \{0,1\}$: 1 if player $i \in P$ is in the squad .
    \item $t_i \in \{0,1\}$: 1 if buy player $i$ (transfer in).
    \item $s_i \in \{0,1\}$: 1 if sell player $i$ (transfer out).
    \item $e \in \{0, 1, 2, \ldots, 15\}$: Number of extra transfers beyond $f$ (penalty).
\end{itemize}

\newpage 

\section{Approach for pre-GW1 and post-GW1 Optimization}

This section presents a solver-independent strategy for FPL optimization.
This decomposition is independent of solver, making it a general algorithmic approach rather than a model-specific formulation.

\subsection{Pre-GW1 Formulation (No Transfers)}

\subsubsection{Step 1: Optimize Starters and Captain}

Decision variables: $y_i, x_i, c_i \in \{0,1\}$ for all $i \in P$.

Solve the following optimization problem:
\begin{equation}
    \max \quad \sum_{i \in P} \text{expected\_points}_i \cdot (x_i + c_i)
\end{equation}
subject to constraints :
\begin{align*}
    \sum_{i \in P} x_i &= 11, \quad \text{(lineup size)} \\
    \sum_{i \in G} x_i &= 1, \quad \text{(1 starting GK)} \\
    3 \leq \sum_{i \in D} x_i &\leq 5, \quad \text{(3-5 starting DEF)} \\
    2 \leq \sum_{i \in M} x_i &\leq 5, \quad \text{(2-5 starting MID)} \\
    1 \leq \sum_{i \in F} x_i &\leq 3, \quad \text{(1-3 starting FWD)} \\
    \sum_{i \in P} c_i &= 1, \quad \text{(exactly 1 captain)} \\
    c_i &\leq x_i \quad \forall i \in P, \quad \text{(captain must be starter)} \\
    u_i + x_i &\leq 1 \quad \forall i \in P, \quad \text{(unavailable players cannot be starter)} \\
    \sum_{i \in P} y_i &= 15, \quad \sum_{i \in G} y_i = 2, \quad \sum_{i \in D} y_i = 5, \\
    \sum_{i \in M} y_i &= 5, \quad \sum_{i \in F} y_i = 3, \quad \text{(squad quotas)} \\
    \sum_{i \in P} p_i y_i &\leq 100, \quad \sum_{i \in P_c} y_i \leq 3 \quad \forall c \in C, \quad \text{(budget and club limits)} \\
    x_i &\leq y_i \quad \forall i \in P \quad \text{(starters must be in squad)}
\end{align*}

Denote the optimal solution from Step 1 as $ x_i^*, y_i^*, c_i^*$.

\subsubsection{Post-hoc: Vice-Captain Assignment}

Let $ S = \{ i \in P \mid x_i^* = 1 \} $ denote the set of starting players.

After solving Step 1, assign the vice-captain as follows: From the starting lineup $S$, exclude the captain and select the player with the highest expected points among the remaining starters.

This ensures the vice-captain captures the best available backup bonus (e.g., for captain no-shows) without altering the optimization in Step 1.

\subsubsection{Step 2: Optimize Bench Composition}

Decision variables: $y_j \in \{0,1\}$ for all $j \in P$.

State variable: $x_i^* \in \{0,1\}$ from Step 1 for all $i \in P$.

Solve the following optimization problem:
\begin{equation}
    \max \quad \sum_{j \in P} \text{expected\_points}_j \cdot y_j
\end{equation}
subject to :
\begin{align*}
    \sum_{i \in P} y_i &= 15 \quad \text{(total squad size)} \\
    \sum_{i \in G} y_i &= 2, \quad \sum_{i \in D} y_i = 5, \quad \sum_{i \in M} y_i = 5, \quad \sum_{i \in F} y_i = 3 \quad \text{(position quotas)} \\
    \sum_{i \in P} p_i y_i &\leq 100, \quad \sum_{i \in P_c} y_i \leq 3 \quad \forall c \in C \quad \text{(budget and club limits)} \\
    y_i &= x_i^* = 1 \quad \forall i \in S \quad \text{(fix starters from Step 1)}
\end{align*}

Denote the optimal solution from step 2 as $y_i^{**}$.

\subsubsection{Post-hoc: Bench Position Assignment}

Assign bench positions as follows to prioritize substitutions of high-expected-point players (bench order determines substitution priority):
\begin{enumerate}
    \item Assign the backup goalkeeper: Place the single bench goalkeeper (the non-starting GK in the squad) into position 1, following the FPL rules.
    \item Sort by expected points: Rank these three benched outfield players from highest to lowest expected points.
    \item Assign to outfield slots: Place the highest-ranked player in position 2 (earliest substitute), the middle-ranked in position 3, and the lowest in position 4 (latest substitute). This ensures the best potential replacement enters first.
\end{enumerate}

This ensures exactly 1 player per bench position, with GK in position 1 and outfield in positions 2-4.

\subsubsection{Output for Pre GW1}
The final output from the Pre-GW1 formulation consists of:
\begin{itemize}
    \item Squad composition: $y_i^{**} \in \{0,1\}$ for all $i \in P$ (15 players).
    \item Starting lineup: $x_i^* \in \{0,1\}$ for all $i \in P$ (11 players).
    \item Captain: $c_i^* \in \{0,1\}$ for all $i \in P$ (1 player).
    \item Vice-captain: assigned post-hoc as per above.
    \item Bench positions: assigned post-hoc as per above.
\end{itemize}

\subsubsection{Initial Condition for State Variables}

After completing GW1 optimization, initialize the state variables for GW2:
\begin{align}
    y_i^0 &\leftarrow y_i^{**} \quad \forall i \in P \quad \text{(GW1 squad becomes initial squad)} \\
    B_{\mathrm{bank}} &= 100 - \sum_{i \in P} p_i y_i^{**} \quad \text{(remaining cash)} \\
    f &= 2 \quad \text{(1 free transfer per week, starts with 1 banked)} \\
    p^0_{\text{buy},i} &= 
    \begin{cases}
        p_i & \text{if } y_i^{**} = 1 \quad \text{(GW1 squad members)} \\
        0 & \text{if } y_i^{**} = 0 \quad \text{(all other players)}
    \end{cases}
    \quad \forall i \in P
\end{align}

These state variables form the complete initial state for the Post-GW1 formulation.

\newpage

\subsection{Post-GW1 Formulation (With Transfers)}

For gameweeks after GW1, the current squad $y_i^0$ is known, and transfers are allowed subject to free transfer limit $f$ and penalty $-4$ points per extra transfer.

\textbf{Selling Price Calculation:}

For each player $i \in P$ in the current squad ($y_i^0 = 1$), the selling price $sp_i$ implements FPL's 50\% profit lock mechanism:
\begin{equation*}
    sp_i = p^0_{\text{buy},i} + 0.5 \cdot \max(0, p_i - p^0_{\text{buy},i})
\end{equation*}
rounded up to nearest 0.1m if profit ($p_i > p^0_{\text{buy},i}$); otherwise $sp_i = p_i$ if loss ($p_i \leq p^0_{\text{buy},i}$).

This ensures:
\begin{itemize}
    \item Players sold at a gain contribute only half their price increase (profit locked at 50\%)
    \item Players sold at a loss contribute their full current price (losses fully realized)
\end{itemize}

\textbf{Budget Dynamics:}

Only new player purchases cost money; retained players are already owned and cost nothing. The budget constraint is derived from the requirement that cash must remain non-negative after transfers.

From the state update (Section 2.2.5):
\begin{equation*}
    B_{\mathrm{bank}}^{\mathrm{new}} = B_{\mathrm{bank}} + \sum_{i \in P} sp_i s_i - \sum_{j \in P} p_j t_j
\end{equation*}

Requiring $B_{\mathrm{bank}}^{\mathrm{new}} \geq 0$ gives:
\begin{equation*}
    B_{\mathrm{bank}} + \sum_{i \in P} sp_i s_i - \sum_{j \in P} p_j t_j \geq 0
\end{equation*}

Rearranging yields the budget constraint:
\begin{equation*}
    \sum_{j \in P} p_j t_j \leq B_{\mathrm{bank}} + \sum_{i \in P} sp_i s_i
\end{equation*}

where $p_j$ is the current market price for new purchases and $sp_i$ is the locked selling price (with 50\% profit lock) for any sold players.

\subsubsection{Step 1: Optimize Starters, Captain, and Transfers}

\textbf{State Variables:} The optimization uses four state variables carried over from the previous gameweek:
\begin{itemize}
    \item $y_i^0 \in \{0,1\}$: Current squad composition (15 players)
    \item $B_{\mathrm{bank}} \in \mathbb{R}^+$: Cash in bank (remaining budget)
    \item $p^0_{\mathrm{buy},i} \in \mathbb{R}^+ \cup \{0\}$: Purchase price for player $i \in P$ (0 if not in squad)
    \item $f \in \{0,1,2,3,4,5\}$: Free transfers available (typically $f=2$ for GW2)
\end{itemize}


Decision variables: $x_i, y_i, c_i, t_i, s_i \in \{0,1\}$ for all $i \in P$, and $e \in \{0, 1, 2, \ldots, 15\}$.

Solve the following optimization problem:
\begin{equation}
    \max \quad \sum_{i \in P} \text{expected\_points}_i \cdot (x_i + c_i) - 4e
\end{equation}
subject to constraints:
\begin{align*}
    \sum_{i \in P} x_i &= 11, \quad \text{(lineup size)} \\
    \sum_{i \in G} x_i &= 1, \quad \text{(1 starting GK)} \\
    3 \leq \sum_{i \in D} x_i &\leq 5, \quad \text{(3-5 starting DEF)} \\
    2 \leq \sum_{i \in M} x_i &\leq 5, \quad \text{(2-5 starting MID)} \\
    1 \leq \sum_{i \in F} x_i &\leq 3, \quad \text{(1-3 starting FWD)} \\
    \sum_{i \in P} c_i &= 1, \quad \text{(exactly 1 captain)} \\
    c_i &\leq x_i \quad \forall i \in P, \quad \text{(captain must be starter)} \\
    u_i + x_i &\leq 1 \quad \forall i \in P, \quad \text{(unavailable players cannot start)} \\
    \sum_{i \in P} y_i &= 15, \quad \sum_{i \in G} y_i = 2, \quad \sum_{i \in D} y_i = 5, \\
    \sum_{i \in M} y_i &= 5, \quad \sum_{i \in F} y_i = 3, \quad \text{(squad quotas)} \\
    \sum_{j \in P} p_j t_j &\leq B_{\mathrm{bank}} + \sum_{i \in P} sp_i s_i, \quad \sum_{i \in P_c} y_i \leq 3 \quad \forall c \in C, \quad \text{(budget and club limits)} \\
    x_i &\leq y_i \quad \forall i \in P \quad \text{(starters must be in squad)} \\
    y_i &= y_i^0 + t_i - s_i \quad \forall i \in P \quad \text{(squad update)} \\
    t_i + s_i &\leq 1 \quad \forall i \in P \quad \text{(no simultaneous buy/sell)} \\
    t_i &\leq 1 - y_i^0 \quad \forall i \in P \quad \text{(buy only non-squad)} \\
    s_i &\leq y_i^0 \quad \forall i \in P \quad \text{(sell only current squad)} \\
    e &\geq \sum_{i \in P} t_i - f \quad \text{(extra transfers)} \\
    \sum_{i \in P} t_i &\leq 15 \quad \text{(logical upper bound on transfers)}
\end{align*}
Denote the optimal solution from Step 1 as $x_i^*, y_i^*, c_i^*, t_i^*, s_i^*, e^*$.

\subsubsection{Post-hoc: Vice-Captain Assignment}

Assign vice-captain as in Pre-GW1.

\subsubsection{Post-hoc: Bench Position Assignment}

Assign $b_{ij}$ greedily as in Pre-GW1.

\subsubsection{Output for Post GW1}
The final output from the Post-GW1 formulation consists of:
\begin{itemize}
    \item Squad composition: $y_i^* \in \{0,1\}$ for all $i \in P$ (15 players, from Step 1).
    \item Starting lineup: $x_i^* \in \{0,1\}$ for all $i \in P$ (11 players, from Step 1).
    \item Captain: $c_i^* \in \{0,1\}$ for all $i \in P$ (1 player, from Step 1).
    \item Vice-captain: assigned post-hoc as per above.
    \item Bench positions: assigned post-hoc as per above.
    \item Transfers made: $t_i^*, s_i^* \in \{0,1\}$ for all $i \in P$ (from Step 1).
    \item Extra transfers penalty: $e^*$ (from Step 1).
\end{itemize}

\subsubsection{State Update for Next Gameweek}

After completing Step 1, update the state variables to carry forward to the next gameweek using per-player cash calculations:

\textbf{Cash Update:}

Compute the new bank balance based on actual selling proceeds and buying costs:
\begin{equation}
    B_{\mathrm{bank}}^{\mathrm{new}} = B_{\mathrm{bank}} + \sum_{i \in P} sp_i s_i^* - \sum_{j \in P} p_j t_j^*
\end{equation}
where:
\begin{itemize}
    \item $\sum_{i \in P} sp_i s_i^*$: Cash from selling players (using locked selling prices)
    \item $\sum_{j \in P} p_j t_j^*$: Cost of buying players (at current market prices)
\end{itemize}

This formulation ensures:
\begin{itemize}
    \item Players sold at a profit contribute $sp_i = p^0_{\text{buy},i} + 0.5(p_i - p^0_{\text{buy},i})$ (50\% profit lock)
    \item Players sold at a loss contribute $sp_i = p_i$ (full current price, loss realized)
    \item The constraint $B_{\mathrm{bank}}^{\mathrm{new}} \geq 0$ is implicitly satisfied by the budget constraints in Step 1
\end{itemize}

\textbf{Purchase Price State Update:}

Update the purchase price history for all players:
\begin{equation}
    \forall i \in P: \quad
    \begin{cases}
        p^0_{\text{buy},i} \leftarrow 0 & \text{if } s_i^* = 1 \quad \text{(sold players - clear history)} \\
        p^0_{\text{buy},i} \leftarrow p_i & \text{if } t_i^* = 1 \quad \text{(bought players - record purchase)} \\
        p^0_{\text{buy},i} \text{ unchanged} & \text{otherwise} \quad \text{(retained/non-squad players)}
    \end{cases}
\end{equation}

\textbf{Squad State Update:}

Update the state variables for the next gameweek:
\begin{itemize}
    \item $y_i^0 \leftarrow y_i^*$ for all $i \in P$ (new squad becomes current squad)
    \item $B_{\mathrm{bank}} \leftarrow B_{\mathrm{bank}}^{\mathrm{new}}$ (updated cash in bank from equation above)
    \item $p^0_{\text{buy},i}$ updated as specified (purchase price tracking)
    \item $f$: Free transfers for next gameweek (1 per week, bankable up to max 5)
\end{itemize}

The decomposition is solver-independent and serves as an algorithmic blueprint for efficient FPL optimization.

\newpage

\subsection{CP-Based Formulations with High-Level Set Abstraction}

This subsection presents equivalent formulations using declarative set-based abstraction, as implemented in MiniZinc constraint programming models. The set-based abstraction maps directly to the binary formulations in Sections 2.1 and 2.2.

\subsubsection{Sets and Parameters}

Let \( P = \{1, \dots, n\} \) be the set of all players.

Let \( C = \{1, \dots, 20\} \) be the set of clubs.

Let \( \mathcal{P} = \{\text{GK}, \text{DEF}, \text{MID}, \text{FWD}\} \) be the set of positions.

For each position \( p \in \mathcal{P} \), let \( P_p \subseteq P \) be the players available in position \( p \).

For each club \( c \in C \), let \( P_c \subseteq P \) be the players in club \( c \).

Let \( e_i \in \mathbb{R}_{\geq 0} \) be the expected points for player \( i \in P \).

Let \( k_i \in \mathbb{R}_{\geq 0} \) be the cost for player \( i \in P \).

Let \( u_i \in \{0,1\} \) indicate if player \( i \in P \) is unavailable.

Let \( U \in \mathbb{R}_{\geq 0} \) be an logical upper bound on the objective.

Let \( L \in \mathbb{R}_{\leq 0} \) be an logical lower bound on the objective.

\subsubsection{Decision Variables}

The following set-based decision variables are common to both Pre-GW1 and Post-GW1 formulations:

Let \( Q \subseteq P \) be the squad set, with \( |Q| = 15 \).

Let \( S \subseteq Q \) be the starters set, with \( |S| = 11 \).

Let \( K \subseteq S \) be the captain set, with \( |K| = 1 \).

For each position \( p \in \mathcal{P} \), let \( Q_p \subseteq P_p \) be the squad partition for \( p \).

For each position \( p \in \mathcal{P} \), let \( S_p \subseteq Q_p \) be the starters partition for \( p \).

\subsubsection{Pre-GW1 Optimization}

This formulation corresponds to the Pre-GW1 problem in Section 2.1.

\textbf{Step 1: Optimize Starters and Captain}

Solve the following optimization problem:
\begin{equation}
    \max_{Q, S, K, \{Q_p, S_p\}_{p \in \mathcal{P}}} \quad z = \sum_{i \in S} e_i + \sum_{i \in K} e_i
\end{equation}
subject to:
\begin{align*}
    0 \leq z &\leq U,
\end{align*}

\textbf{Cardinalities:}
\begin{align*}
    &|Q| = 15, \quad |S| = 11, \quad |K| = 1, \\
    &|Q_{\text{GK}}| = 2, \quad |Q_{\text{DEF}}| = 5, \quad |Q_{\text{MID}}| = 5, \quad |Q_{\text{FWD}}| = 3, \\
    &|S_{\text{GK}}| = 1, \quad 3 \leq |S_{\text{DEF}}| \leq 5, \quad 2 \leq |S_{\text{MID}}| \leq 5, \quad 1 \leq |S_{\text{FWD}}| \leq 3,
\end{align*}

\textbf{Partitions:}
\begin{align*}
    &Q = \bigcup_{p \in \mathcal{P}} Q_p, \quad \forall p_1 \neq p_2 \in \mathcal{P} : \, Q_{p_1} \cap Q_{p_2} = \emptyset, \\
    &S = \bigcup_{p \in \mathcal{P}} S_p, \quad \forall p_1 \neq p_2 \in \mathcal{P} : \, S_{p_1} \cap S_{p_2} = \emptyset, \\
    &\forall p \in \mathcal{P} : \, Q_p \subseteq P_p, \, S_p \subseteq Q_p,
\end{align*}

\textbf{Budget:}
\begin{align*}
    \sum_{i \in Q} k_i &\leq 100,
\end{align*}

\textbf{Club Limits:}
\begin{align*}
    \forall c \in C &: \, |Q \cap P_c| \leq 3,
\end{align*}

\textbf{Availability:}
\begin{align*}
    \forall i \in P &: \, (u_i = 1) \implies (i \notin S),
\end{align*}

\textbf{Subsets:}
\begin{align*}
    S &\subseteq Q, \quad K \subseteq S.
\end{align*}

Denote the optimal solution from Step 1 as \( Q^*, S^*, K^*, \{Q_p^*, S_p^*\}_{p \in \mathcal{P}} \).

\textbf{Step 2: Optimize Bench}

Fix \( S^* \) and \( \{S_p^* \mid p \in \mathcal{P}\} \) from Step 1's optimal solution. Then solve:
\begin{equation}
    \max_{Q, \{Q_p\}_{p \in \mathcal{P}}} \quad z = \sum_{i \in Q} e_i
\end{equation}
subject to:
\begin{align*}
    0 \leq z &\leq U, \\
    S &= S^*, \quad \forall p \in \mathcal{P} : \, S_p = S_p^*, \\
\end{align*}
and all other constraints from Step 1.

Denote the optimal solution from Step 2 as \( Q^{**}, \{Q_p^{**}\}_{p \in \mathcal{P}} \).

\textbf{Post-hoc Assignments:}

After solving steps 1 and 2, perform the same post-hoc assignments as in Section 2.1 (vice-captain and bench position assignments).

\subsubsection{Post-GW1 Optimization with Transfers}

This formulation corresponds to the Post-GW1 problem in Section 2.2.

The key design choice is using explicit binary variables \(t_i, s_i\) for transfers (rather than set differences) to avoid propagation issues, while maintaining high-level set abstractions for squad/starters.

\textbf{Additional State Variables:}

For Post-GW1, the following state variables from the previous gameweek are required:

Let \( y^0_i \in \{0,1\} \) indicate if player \( i \in P \) is in the current squad.

Let \( sp_i \in \mathbb{R}_{\geq 0} \) be the selling price for player \( i \in P \) (scaled by 10, computed via 50\% profit lock).

Let \( B_{\text{bank}} \in \mathbb{R}_{\geq 0} \) be the cash in bank (scaled by 10).

Let \( f \in \{0,1,2,3,4,5\} \) be the number of free transfers available.

\textbf{Additional Decision Variables:}

For each player \( i \in P \), let \( t_i \in \{0,1\} \) indicate transfer in (buy player \(i\)).

For each player \( i \in P \), let \( s_i \in \{0,1\} \) indicate transfer out (sell player \(i\)).

Let \( e \in \{0,1,\dots,15\} \) be the number of extra transfers beyond \(f\).

For each player \( i \in P \), let \( y_i \in \{0,1\} \) indicate if player \(i\) is in the new squad (channeled with set \(Q\)).

\textbf{Optimization Formulation:}

Solve the following optimization problem:
\begin{equation}
    \max_{Q, S, K, \{Q_p, S_p\}_{p \in \mathcal{P}}, \{y_i, t_i, s_i\}_{i \in P}, e} \quad z = \sum_{i \in S} e_i + \sum_{i \in K} e_i - 40 \cdot e
\end{equation}
subject to:
\begin{align*}
    L \leq z &\leq U \quad \text{(precalculated logical bound)},
\end{align*}

\textbf{Transfer Logic:}
\begin{align*}
    &\forall i \in P : \, y_i = y^0_i + t_i - s_i \quad \text{(squad update)}, \\
    &\forall i \in P : \, t_i + s_i \leq 1 \quad \text{(no buy and sell same player)}, \\
    &\forall i \in P : \, y^0_i + t_i \leq 1 \quad \text{(no buy if owned)}, \\
    &\forall i \in P : \, s_i \leq y^0_i \quad \text{(only sell owned)}, \\
    &e \geq \sum_{i \in P} t_i - f \quad \text{(extra transfers penalty)}, \\
    &\sum_{i \in P} t_i \leq 15 ,
\end{align*}

\textbf{Budget with Transfers:}
\begin{align*}
    \sum_{i \in P} k_i \cdot t_i &\leq B_{\text{bank}} + \sum_{i \in P} sp_i \cdot s_i,
\end{align*}

\textbf{Channeling and Subsets:}
\begin{align*}
    \forall i \in P : \, (y_i = 1) &\iff (i \in Q), \\
    S &\subseteq Q, \quad K \subseteq S,
\end{align*}

\textbf{Cardinalities:}
\begin{align*}
    \sum_{i \in P} y_i &= 15, \quad |S| = 11, \quad |K| = 1, \\
    |Q_{\text{GK}}| &= 2, \quad |Q_{\text{DEF}}| = 5, \quad |Q_{\text{MID}}| = 5, \quad |Q_{\text{FWD}}| = 3, \\
    |S_{\text{GK}}| &= 1, \quad 3 \leq |S_{\text{DEF}}| \leq 5, \quad 2 \leq |S_{\text{MID}}| \leq 5, \quad 1 \leq |S_{\text{FWD}}| \leq 3,
\end{align*}

\textbf{Partitions:}
\begin{align*}
    &Q = \bigcup_{p \in \mathcal{P}} Q_p, \quad \forall p_1 \neq p_2 \in \mathcal{P} : \, Q_{p_1} \cap Q_{p_2} = \emptyset, \\
    &S = \bigcup_{p \in \mathcal{P}} S_p, \quad \forall p_1 \neq p_2 \in \mathcal{P} : \, S_{p_1} \cap S_{p_2} = \emptyset, \\
    & \forall p \in \mathcal{P} : \, Q_p \subseteq P_p, \, S_p \subseteq Q_p,
\end{align*}

\textbf{Club Limits and Availability:}
\begin{align*}
    &\forall c \in C : \, |Q \cap P_c| \leq 3, \\
    &\forall i \in P : \, u_i = 1 \implies i \notin S.
\end{align*}

\textbf{Post-hoc Assignments:}

After solving the optimization, perform the same post-hoc assignments as in Section 2.2 (vice-captain and bench position assignments). State updates follow Section 2.2.5.

\end{document}
